\gfcommand{let}{evaluation plus attachment}
\index{let}

\gfsyntax{
  let \ARG{\indconst} to \ALT dar [ \ARG{rep} ] \ARG{term};
}

\gfdescription{
  This command evaluates the \ARG{term} using the mixed evaluation mechanism
  of the {\tt eval} command.
  If the evaluation returns a {\HG} representation for \ARG{term}, then it is
  attached to  \ARG{indconst} with representation function \ARG{rep}. Then the 
  equality of \ARG{indconst} with \ARG{term} is asserted as the next line in the
  proof, otherwise an error message is given. If \ARG{rep} is not specified the
  representation is the default representation. If this does not happen {\GF}
  outputs an error message.
}

\gfrecap{
This command evaluates the `term' using the mixed evaluation mechanism
of the {\tt eval} command.
If the evaluation returns a HGKM representation for `term', then it is
attached to  `indconst' with representation function `rep'. Then the 
equality of `indconst' with `term' is asserted as the next line in the
proof, otherwise an error message is given. If `rep' is not specified the
representation is the default representation. If this does not happen GETFOL
outputs an error message.
}

\gfexample+
   ***** declare indconst a b c;
   ***** attach b to b;
   b attached to 'b
   ***** attach c to c;
   c attached to 'c
   ***** declare funconst h 2;
   ***** DEFLAM h(x y) (QUOTE d);
   ***** attach h to h;
   h attached to h
   ***** let a dar h(b c);
   a attached to 'd
   a is the preferred name of d
   1   a = h(b,c)     
+
